Exploring the Utility of Machine Learning Across Varied Data Formats

### 2024-05-23 homework2/paper/paper.tex
modify second chapter

\begin{document}
%
\title{Exploring the Utility of Machine Learning Across Varied Data Formats}
%
\author{Cărămidă Iustina-Andreea - 332CA}
%
\institute{Faculty of Automatic Control and Computer Science \\
University Politehnica of Bucharest \\
\email{iustina.caramida@stud.acs.upb.ro}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    This study investigates the applicability of machine learning techniques on 
    diverse datasets. We explore the effectiveness of two algorithms, Linear 
    Regression and Multi-Layered Perceptron (MLP), on predicting both health 
    outcomes and financial well-being. Specifically, we utilize a stroke 
    prediction dataset to assess the model\'s ability to identify individuals at 
    risk of stroke. Additionally, we employ a salary prediction dataset to 
    evaluate the model\'s capacity to classify individuals earning above a 
    specific income threshold (e.g., \$50,000 per year). Through comparative 
    analysis, this research aims to elucidate the strengths and limitations of 
    each algorithm when applied to these contrasting data types, offering 
    insights into their suitability for various prediction tasks. Furthermore, 
    we present a framework for data analysis, outlining essential steps for data
    cleaning, exploration, and preparation, which can be applied to enhance the 
    effectiveness of machine learning models across diverse datasets.

@@

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{3.5cm}||>{\centering\arraybackslash}m{2cm}|>{\centering\arraybackslash}m{3.5cm}||>{\centering\arraybackslash}m{2.5cm}| }
        \hline
        \multicolumn{4}{|c|}{List of all attributes in the Stroke Prediction dataset} \\
        \hline
        Attribute name & Type & Details & Possible values \\
        \hline
        mean\_blood\_sugar\_ level & numeric & The average value of blood glucose throughout the duration observation of the subject & \\
        \hline
        cardiovascular\_issues & categorical & Whether or not the subject has a medical history cardiovascular & 0, 1 \\
        \hline
        job\_category & categorical & The field in which the person works & child, entrepreneurial, N\_work\_history, private\_sector, public\_sector \\
        \hline
        body\_mass\_indicator & numeric & Body mass index, which indicates if the person is underweight, within limits normal, overweight or obese & \\
        \hline
        sex & categorical & The gender of the person & F, M\\
        \hline
        tobacco\_usage & categorical & Current or past smoker indicator &  ex-smoker, smoker, non-smoker\\
        \hline
        high\_blood\_pressure & categorical & Binary attribute indicating whether a person suffer from high blood pressure or not & 0, 1 \\
        \hline
        married & categorical & Binary attribute indicating whether the person a ever been married & Y, N \\
        \hline
        living\_area & categorical & The type of area where he lived most of his life & City, Countryside \\
        \hline
        years\_old & numeric & The person's age in years & \\
        \hline
        chaotic\_sleep & categorical & Binary attribute for a sleep program irregular & 0, 1 \\
        \hline
        analysis\_results & numeric & The results of medical analyzes of the person, which may include various measurements and indicators relevant to her health & \\
        \hline
        biological\_age\_index & numeric & An index that estimates the biological age of a person based on different factors such as lifestyle, health status, measured in an unknown unit & \\
        \hline
        cerebrovascular\_ accident & categorical & Binary indicator indicating whether the person a had a stroke or not & 0, 1 \\
        \hline
        \caption{Stroke Prediction Attributes}
        \label{tab:stroke_attributes_description}
   \end{longtable}
\end{center}

\subsection{Exploration of Attribute Types and Value Ranges}
Prior to applyinga machine learning model to a dataset, a crucial step involves
in identifying the types of attributes (features) present and their corresponding
values ranges. This anaysis is essential for selecting appropriate algorithms 
and ensuring optimalmodel performance. In the following pharagraphs we will
describe three primary attribute types.
\begin{itemize}
    \item \textit{Continuous Numeric Attributes:}
    These attributes possess numerical values that can theoretically take on any 
    value within a specific range. Examples might include: age, weight, temperature etc.
    \item \textit{Discrete Nominal Attributes:}
    These attributes represent categorical data with distinct, non-ordered values. 
    Examples include days of the week (Monday, Tuesday, etc.) or types of diseases 
    (cancer, diabetes, etc.).
    \item \textit{Ordinal Attributes:}
    These attributes represent categorical data with values that exhibit an inherent 
    order. However, the difference between consecutive values may not be interpretable 
    in terms of a consistent unit.  Examples include customer satisfaction ratings 
    (1-star, 2-star, etc.) or movie ratings (G, PG, PG-13, etc.). In ordinal 
    attributes, the numerical value itself might not be as important as the relative 
    order it represents.
\end{itemize}

Using the \textit{analysis\_attributes.py} script, we can identify the 
Continuous Numeric Attributes and Discrete Nominal Attributes in a specific 
dataset. The script will output statistics that can be showed in Tables 
\ref{tab:continuous_numeric_attributes_salary} and 
\ref{tab:continuous_numeric_attributes_stroke} for numeric attributes and
Table \ref{tab:discrete_nominal_attributes_salary} and 
\ref{tab:discrete_nominal_attributes_stroke} for nominal attributes.

@@

An initial inspection of the data reveals that there are missing attributes in 
both the Salary Prediction and Stroke Prediction datasets. The Salary Prediction 
dataset is missing the \textit{'hpw'} attribute, while the Stroke Prediction dataset is
missing two attributes: \textit{'body\_mass\_indicator'} and 
\textit{'analysis\_results'}.

@@

Also, in Figure \ref{fig:boxplot_example_stroke} we can see a boxplot for the
\textit{body\_mass\_indicator} attribute in the Stroke Prediction dataset.
As described above, the boxplot provides a visual representation of the data's
distribution, highlighting key statistical measures such as the median, quartiles,
and potential outliers. These information is also presented in the tables
\ref{tab:continuous_numeric_attributes_salary} and
\ref{tab:continuous_numeric_attributes_stroke}. One of the main insights that
can be derived from the boxplot is the presence of outliers, which are data points
that lie significantly outside the range of the rest of the data. Outliers can
have a significant impact on the performance of machine learning models, and
identifying and handling them appropriately is an essential step in the data
preprocessing process.

@@

From the Discret Nominal Attributes tables we can see that each dataset contains
only one attribute with missing values. The Salary Prediction dataset is missing
the \textit{gender} attribute, while the Stroke Prediction dataset is missing the
\textit{married} attribute. Also, the number of unique values for each attribute
descibes the diversity of the data. For example, the \textit{country} attribute
in the Salary Prediction dataset has 41 unique values, indicating that the data
contains information from 41 different countries.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/distribution_edu_Salary_Prediction.png}
    \caption{Distribution of the \textit{edu} attribute in the Salary Prediction dataset}
    \label{fig:distribution_example_salary}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/distribution_tobacco_usage_Stroke_Prediction.png}
    \caption{Distribution of the \textit{tobacco\_usage} attribute in the Stroke Prediction dataset}
    \label{fig:distribution_example_stroke}
\end{figure}

In the \textit{'plots'} folder at the root of the project directory we can find
the distribution of each class in the datasets. The name of each plot starts
with \textit{'distribution\_'}. Here is an example of the distribution of the
\textit{edu} attribute in the Salary Prediction dataset in Figure
\ref{fig:distribution_example_salary}. Also, the distribution of the
\textit{tobacco\_usage} attribute in the Stroke Prediction dataset is shown in
Figure \ref{fig:distribution_example_stroke}.

@@

The correlation matrix and Cramér's V matrix provide valuable insights into the
relationships between attributes in the datasets.By examining these matrices, we
can see that Figure \ref{fig:correlation_matrix_example_salary} the \textit{prod}
attribute is highly correlated with the \textit{gain} attribute, while the
\textit{years} attribute is negatively correlated with the \textit{fnl} attribute.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/correlation_matrix_SalaryPrediction_full.png}
    \caption{Correlation matrix for the Salary Prediction dataset}
    \label{fig:correlation_matrix_example_salary}
\end{figure}

@@

% ---- Bibliography ----
\begin{thebibliography}{8}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Correlation}{Wikipedia - Correlation}
    \bibitem{}
    \href{https://thinkingneuron.com/how-to-measure-the-correlation-between-two-categorical-variables-in-python/}{Thinking Neuron - How to Measure the Correlation Between Two Categorical Variables in Python}
    \bibitem{}
    \href{https://www.stratascratch.com/blog/chi-square-test-in-python-a-technical-guide/}{StrataScratch - Chi-Square Test in Python: A Technical Guide}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Cram\%C3\%A9r\%27s\_V}{Wikipedia - Cramér's V}
    \bibitem{}

>>>>

\begin{document}
%
\title{Exploring the Utility of Machine Learning Across Varied Data Formats}
%
\author{Cărămidă Iustina-Andreea - 332CA}
%
\institute{Faculty of Automatic Control and Computer Science \\
University Politehnica of Bucharest \\
\email{iustina.caramida@stud.acs.upb.ro}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    This study investigates the applicability of machine learning techniques on 
    diverse datasets. We explore the effectiveness of two algorithms, Linear 
    Regression and Multi-Layered Perceptron (MLP), on predicting both health 
    outcomes and financial well-being. Specifically, we utilize a stroke 
    prediction dataset to assess the model's ability to identify individuals at 
    risk of stroke. Additionally, we employ a salary prediction dataset to 
    evaluate the model's capacity to classify individuals earning above a 
    specific income threshold (e.g., \$50,000 per year). Through comparative 
    analysis, this research aims to elucidate the strengths and limitations of 
    each algorithm when applied to these contrasting data types, offering 
    insights into their suitability for various prediction tasks. Furthermore, 
    we present a framework for data analysis, outlining essential steps for data
    cleaning, exploration, and preparation, which can be applied to enhance the 
    effectiveness of machine learning models across diverse datasets.

@@

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{3.5cm}||>{\centering\arraybackslash}m{2cm}|>{\centering\arraybackslash}m{3.5cm}|>{\centering\arraybackslash}m{2.5cm}| }
        \hline
        \multicolumn{4}{|c|}{List of all attributes in the Stroke Prediction dataset} \\
        \hline
        Attribute name & Type & Details & Possible values \\
        \hline
        mean\_blood\_sugar\_ level & numeric & The average value of blood glucose throughout the duration observation of the subject & \\
        \hline
        cardiovascular\_issues & categorical & Whether or not the subject has a medical history cardiovascular & 0, 1 \\
        \hline
        job\_category & categorical & The field in which the person works & child, entrepreneurial, N\_work\_history, private\_sector, public\_sector \\
        \hline
        body\_mass\_indicator & numeric & Body mass index, which indicates if the person is underweight, within limits normal, overweight or obese & \\
        \hline
        sex & categorical & The gender of the person & F, M\\
        \hline
        tobacco\_usage & categorical & Current or past smoker indicator &  ex-smoker, smoker, non-smoker\\
        \hline
        high\_blood\_pressure & categorical & Binary attribute indicating whether a person suffer from high blood pressure or not & 0, 1 \\
        \hline
        married & categorical & Binary attribute indicating whether the person a ever been married & Y, N \\
        \hline
        living\_area & categorical & The type of area where he lived most of his life & City, Countryside \\
        \hline
        years\_old & numeric & The person's age in years & \\
        \hline
        chaotic\_sleep & categorical & Binary attribute for a sleep program irregular & 0, 1 \\
        \hline
        analysis\_results & numeric & The results of medical analyzes of the person, which may include various measurements and indicators relevant to her health & \\
        \hline
        biological\_age\_index & numeric & An index that estimates the biological age of a person based on different factors such as lifestyle, health status, measured in an unknown unit & \\
        \hline
        cerebrovascular\_ accident & categorical & Binary indicator indicating whether the person a had a stroke or not & 0, 1 \\
        \hline
        \caption{Stroke Prediction Attributes}
        \label{tab:stroke_attributes_description}
   \end{longtable}
\end{center}

\subsection{Exploration of Attribute Types and Value Ranges}
Prior to applying a machine learning model to a dataset, a crucial step involves
in identifying the types of attributes (features) present and their corresponding
values ranges. This anaysis is essential for selecting appropriate algorithms 
and ensuring optimal model performance. In the following pharagraphs we will
describe three primary attribute types.
\begin{itemize}
    \item \textit{Continuous Numeric Attributes:}
    These attributes possess numerical values that can theoretically take on any 
    value within a specific range. Examples might include: age, weight, temperature etc.
    \item \textit{Discrete Nominal Attributes:}
    These attributes represent categorical data with distinct, non-ordered values. 
    Examples include days of the week (Monday, Tuesday, etc.) or types of diseases 
    (cancer, diabetes, etc.).
    \item \textit{Ordinal Attributes:}
    These attributes represent categorical data with values that exhibit an inherent 
    order. However, the difference between consecutive values may not be interpretable 
    in terms of a consistent unit.  Examples include customer satisfaction ratings 
    (1-star, 2-star, etc.) or movie ratings (G, PG, PG-13, etc.). In ordinal 
    attributes, the numerical value itself might not be as important as the relative 
    order it represents.
\end{itemize}

Using the \textit{analysis\_attributes.py} script, we can identify the 
Continuous Numeric Attributes and Discrete Nominal Attributes in the  
datasets. The script will output statistics that can be showed in Tables 
\ref{tab:continuous_numeric_attributes_salary} and 
\ref{tab:continuous_numeric_attributes_stroke} for numeric attributes and
Table \ref{tab:discrete_nominal_attributes_salary} and 
\ref{tab:discrete_nominal_attributes_stroke} for discrete attributes.

@@

An initial inspection of the data reveals that there are missing attributes in 
both the Salary Prediction and Stroke Prediction datasets. In the Salary Prediction 
datase the \textit{'hpw'} attribute is missing, while int the Stroke Prediction dataset
two attributes are missing: \textit{'body\_mass\_indicator'} and 
\textit{'analysis\_results'}.

@@

Also, in Figure \ref{fig:boxplot_example_stroke} we can see a boxplot for the
\textit{body\_mass\_indicator} attribute in the Stroke Prediction dataset.
As described above, the boxplot provides a visual representation of the data's
distribution, highlighting key statistical measures such as the median, quartiles,
and potential outliers. These information is also presented in Tables
\ref{tab:continuous_numeric_attributes_salary} and
\ref{tab:continuous_numeric_attributes_stroke}. One of the main insights that
can be derived from the boxplot is the presence of outliers, which are data points
that lie significantly outside the range of the rest of the data. Outliers can
have a significant impact on the performance of machine learning models, and
identifying and handling them appropriately is an essential step in the data
preprocessing process.

@@

From the Discret Nominal Attributes tables (Tables \ref{tab:discrete_nominal_attributes_salary} and \ref{tab:discrete_nominal_attributes_stroke})
we can see that each dataset contains
only one attribute with missing values. In the Salary Prediction dataset 
the \textit{gender} attribute is missing, while the Stroke Prediction dataset the
\textit{married} attribute is missing. Also, the number of unique values for each attribute
descibes the diversity of the data. For example, the \textit{country} attribute
in the Salary Prediction dataset has 41 unique values, indicating that the data
contains information from 41 different countries.

@@

In Figures \ref{fig:distribution_salary} and \ref{fig:distribution_stroke} we can
see the distribution of each class in the datasets. The class distributions provide
insights into the balance of the data and can help guide the selection of appropriate
strategies for handling imbalanced classes. For example, in the Stroke Prediction
dataset, the \textit{cerebrovascular\_accident} class is highly imbalanced, with
a significantly higher number of negative instances compared to positive instances.
This imbalance can impact the model's ability to learn patterns from the minority
class and may require resampling techniques or cost-sensitive learning to address.
On the other hand, the Salary Prediction dataset exhibits a more balanced distribution
of the \textit{money} class, which may require less intervention to handle class
imbalance.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/distribution_SalaryPrediction.png}
    \caption{Distribution of the \textit{money} class in the Salary Prediction dataset}
    \label{fig:distribution_salary}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/distribution_AVC.png}
    \caption{Distribution of the \textit{cerebrovascular\_accident} class in the Stroke Prediction dataset}
    \label{fig:distribution_stroke}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/correlation_matrix_SalaryPrediction_full.png}
    \caption{Correlation matrix for the Salary Prediction dataset}
    \label{fig:correlation_matrix_example_salary}
\end{figure}
\pagebreak
The correlation matrix and Cramér's V matrix provide valuable insights into the
relationships between attributes in the datasets.By examining these matrices, we
can see that Figure \ref{fig:correlation_matrix_example_salary} the \textit{prod}
attribute is highly correlated with the \textit{gain} attribute, while the
\textit{years} attribute is negatively correlated with the \textit{fnl} attribute.

@@

@@

% ---- Bibliography ----
\begin{thebibliography}{8}
    \bibitem{}
    \href{https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.html}{Pandas - DataFrame}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Correlation}{Wikipedia - Correlation}
    \bibitem{}
    \href{https://thinkingneuron.com/how-to-measure-the-correlation-between-two-categorical-variables-in-python/}{Thinking Neuron - How to Measure the Correlation Between Two Categorical Variables in Python}
    \bibitem{}
    \href{https://www.stratascratch.com/blog/chi-square-test-in-python-a-technical-guide/}{StrataScratch - Chi-Square Test in Python: A Technical Guide}
    \bibitem{}
    \href{https://en.wikipedia.org/wiki/Cram\%C3\%A9r\%27s\_V}{Wikipedia - Cramér's V}
    \bibitem{}
